The Talking Heads Experiment

### 2015-04-27 chap1/chap1.tex
fixed index entries and extra papers, added proofreaders and typesetters

For centuries, philosophers, linguists, psychologists and 
neuroscientists have been groping with the amazing capacities
of the mind. By necessity, they have been doing this through
thought experiments or by observing human behaviour and
brain anatomy. Although this has generated a wealth
of insights,\footnote{Attempts are made to bring together the insights from various
disciplines to establish a true ``cognitive science''. 
\cite{Osherton:1995} contains an introduction into some of the main 
research trends in this very diverse scientific field. 
See also \cite{Luger:1994}. The work reported in the present book can be 
classified as theoretical
cognitive science because I try to formulate and test
the operational adequacy of possible models for the 
origins of language and meaning but do not claim nor
give any evidence that these models are also valid for
human cognition, just as the study of aerodynamics and 
aircraft design may help to understand how birds can 
fly but is more generic than its biological 
implementations.}
everyone involved in this research must surely 
agree that we are still lacking adequate models, 
particularly for higher order cognitive functions like language
processing, and definitely for understanding how such 
functions may have arisen. 
To discover or test such models, it therefore
remains useful to do experiments with artificial systems.
We can build robots that receive sensory inputs
through a camera or other sensors, give them computational 
power and memory, and empower them 
for action in the world by adding actuators. Given such a 
set-up, we can precisely examine the operational adequacy 
of an hypothesis. For example, if someone proposes
a process for segmenting images, we can test this process 
by capturing streams of images through a camera and 
see whether the process is indeed capable of performing 
segmentation. When someone proposes a process 
for parsing sentences, we 
can implement this process and confront it with a series of 
example sentences to measure success and failure. 
How else can we test the operational adequacy of a proposed
cognitive model, given the enormous complexity involved? 
Of course, building an artificial system in no way proves
that the principles that were used to construct it are valid
for natural systems. But it is an enormously valuable 
source of insight for approaching the extraordinarily 
complex phenomena observed in human cognition.

@@

A fourth radical idea concerns the nature and origins of 
grammar. Rather than invoking the need of a highly 
specialised genetically determined language 
organ,\footnote{
As strongly argued by Chomsky in various writings. 
See for example \cite{Chomsky:1968}. Even though Chomsky argues for 
an innate language acquisition device he has expressed
scepticism about evolution by natural selection
as an explanation for the 
origin of this device (and therefore of language). 
A genetic theory of language evolution 
has been suggested by \cite{Pinker:1994}.} 
I believe that grammar spontaneously arises when 
generic capabilities to categorise reality, store past
events in terms of abstract schemas, remember associations
between events, etc., reach a critical level and are
applied to language itself. These capabilities are 
relevant across many different cognitive domains. 
In order to store linguistic
experiences, human memories spontaneously structure them, thus 
introducing abstract schemas, internal categories, and 
roles that substructures can play in schemas.
These organisational elements then become externalised. 
Categories are marked by enriching
the form of words, schema boundaries are marked by 
imposing patterns on the expression of a schema, 
roles are marked by assigning
them to specific positions in a 
pattern. This externalisation 
increases the reliability in communication because
it reduces ambiguities and supplies additional context. It 
also aids the stable transmission of the language from one
generation to the next because the language learner
gets additional hints to guess the meaning and function of unknown
words and constructions. 
Once such structuring devices are in place, they help 
to increase the expressive power of the language. The 
complexity of possible language interactions can increase, 
and words or word groups which had multiple usages
can become specialised for particular semantic functions.

@@

In 1996, a computer program called Deep Blue
defeated the reigning world champion Kasparov in a game of 
chess.\footnote{See: \cite{Newborn:1996}.}
Kasparov was astonished
and depressed, and claimed this was a defeat of humans in 
the 
race against machines. But was he actually beaten by {\itshape artificial}
intelligence? Not really. A team of engineers and scientists
from Carnegie Mellon University and from the IBM Watson 
Research Center had been working for ten years to program 
vast amounts of chess knowledge, invented by
human experts, into Deep Blue. They had built extremely 
sophisticated dedicated computing hardware to
apply this knowledge at a blinding speed. So, in beating Kasparov, 
other humans were the clever ones, not machines.

@@

Secondly, the Talking Heads experiment turned out to be 
an ideal learning environment for raising philosophical
issues. Children and adults alike
started to ask questions about the nature of
meaning, the relation between language
and reality, the mind-body problem, 
the origins and evolution of language, 
consciousness, social identity, and so on. They 
start to play language games among themselves and some of
them reported profound changes in the way they 
think about language. If these reflections create 
greater tolerance towards other languages and 
the conceptualisations of the world
they implicitly embody, then I consider the
Talking Heads experiment of great societal value, irrespective of
the scientific and engineering breakthroughs
the project has generated.

>>>>

For centuries, philosophers, linguists, psychologists and 
neuroscientists have been groping with the amazing capacities
of the mind. By necessity, they have been doing this through
thought experiments or by observing human behaviour and
brain anatomy. Although this has generated a wealth
of insights,\footnote{Attempts are made to bring together the insights from various
disciplines to establish a true ``cognitive science''. 
\cite{Osherton:1995} contains an introduction into some of the main 
research trends in this very diverse scientific field. 
See also \cite{Luger:1994}. The work reported in the present book can be 
classified as theoretical
cognitive science because I try to formulate and test
the operational adequacy of possible models for the 
origins of language and meaning but do not claim nor
give any evidence that these models are also valid for
human cognition, just as the study of aerodynamics and 
aircraft design may help to understand how birds can 
fly but is more generic than its biological 
implementations.}
everyone involved in this research must surely 
agree that we are still lacking adequate models, 
particularly for higher order cognitive functions like language
processing, and definitely for understanding how such 
functions may have arisen. 
To discover or test such models, it therefore
remains useful to do experiments with artificial systems.
We can build robots that receive sensory inputs
through a camera or other sensors, give them computational 
power and memory, and empower them 
for action in the world by adding actuators. Given such a 
set-up, we can precisely examine the operational adequacy 
of a hypothesis. For example, if someone proposes
a process for segmenting images, we can test this process 
by capturing streams of images through a camera and 
see whether the process is indeed capable of performing 
segmentation. When someone proposes a process 
for parsing sentences, we 
can implement this process and confront it with a series of 
example sentences to measure success and failure. 
How else can we test the operational adequacy of a proposed
cognitive model, given the enormous complexity involved? 
Of course, building an artificial system in no way proves
that the principles that were used to construct it are valid
for natural systems. But it is an enormously valuable 
source of insight for approaching the extraordinarily 
complex phenomena observed in human cognition.

@@

A fourth radical idea concerns the nature and origins of 
grammar. Rather than invoking the need of a highly 
specialised genetically determined language 
organ,\footnote{As strongly argued by Chomsky in various writings, for example \cite{Chomsky:1968}. Even though Chomsky argues for 
an innate language acquisition device he has expressed
scepticism about evolution by natural selection
as an explanation for the 
origin of this device (and therefore of language). 
A genetic theory of language evolution 
has been suggested by \cite{Pinker:1994}.} 
I believe that grammar spontaneously arises when 
generic capabilities to categorise reality, store past
events in terms of abstract schemas, remember associations
between events, etc., reach a critical level and are
applied to language itself. These capabilities are 
relevant across many different cognitive domains. 
In order to store linguistic
experiences, human memories spontaneously structure them, thus 
introducing abstract schemas, internal categories, and 
roles that substructures can play in schemas.
These organisational elements then become externalised. 
Categories are marked by enriching
the form of words, schema boundaries are marked by 
imposing patterns on the expression of a schema, 
roles are marked by assigning
them to specific positions in a 
pattern. This externalisation 
increases the reliability in communication because
it reduces ambiguities and supplies additional context. It 
also aids the stable transmission of the language from one
generation to the next because the language learner
gets additional hints to guess the meaning and function of unknown
words and constructions. 
Once such structuring devices are in place, they help 
to increase the expressive power of the language. The 
complexity of possible language interactions can increase, 
and words or word groups which had multiple usages
can become specialised for particular semantic functions.

@@

In 1996, a computer program called Deep Blue
defeated the reigning world champion Kasparov in a game of 
chess \citep{Newborn:1996}.
Kasparov was astonished
and depressed, and claimed this was a defeat of humans in 
the 
race against machines. But was he actually beaten by {\itshape artificial}
intelligence? Not really. A team of engineers and scientists
from Carnegie Mellon University and from the IBM Watson 
Research Center had been working for ten years to program 
vast amounts of chess knowledge, invented by
human experts, into Deep Blue. They had built extremely 
sophisticated dedicated computing hardware to
apply this knowledge at a blinding speed. So, in beating Kasparov, 
other humans were the clever ones, not machines.

@@

Secondly, the Talking Heads experiment turned out to be 
an ideal learning environment for raising philosophical
issues. Children and adults alike
started to ask questions about the nature of
meaning, the relation between language
and reality, the mind-body problem, 
the origins and evolution of language, 
consciousness, social identity, and so on. They 
started to play language games among themselves and some of
them reported profound changes in the way they 
think about language. If these reflections create 
greater tolerance towards other languages and 
the conceptualisations of the world
they implicitly embody, then I consider the
Talking Heads experiment of great societal value, irrespective of
the scientific and engineering breakthroughs
the project has generated.
